How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?

### 2024-02-13 outputs/paper/paper.qmd
added intro and abstract and updated llm usage and readme

---
title: "How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?"
author: 
  - Adrian Ly
  - Sakhil Goel
  - Hannah Yu
thanks: "Code and data are available at: https://github.com/hannahyu07/US-Covid-Analysis.git"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

@fig-GHS allows us to visualize the Global Health Security (GHS) Index which is designed to assess a country's capabilities to prevent, detect, and respond to significant infectious disease outbreaks. A higher Global Health Security (GHS) Index score indicates a stronger health system capacity for pandemic readiness. The GHS Index assesses the readiness of countries for pandemics and other significant biological threat emergencies, using a comprehensive framework that includes measurements grouped into six broad categories: prevention, detection, response, health system capacity, compliance with international norms, and risk environment. Higher scores suggest that a country has a more robust health system capacity, with better capabilities to prevent, detect, respond to, and mitigate the spread of an epidemic.

@@

>>>>

---
title: "How did local health infrastructure and socio-political factors within different states and counties in the United States affect the disparities in COVID-19 outcomes, and what lessons can be learned for more targeted public health preparedness and response strategies in future pandemics?"
author: 
  - Adrian Ly
  - Sakhil Goel
  - Hannah Yu
thanks: "Code and data are available at: https://github.com/hannahyu07/US-Covid-Analysis.git"
date: today
date-format: long
abstract: "In this research, we explored the complex interplay between local health infrastructure, socio-political factors, and COVID-19 outcomes across various U.S. states and counties to understand disparities in the pandemic's impact. Our findings reveal that political polarization and healthcare access disparities significantly influenced adherence to health guidelines and the effectiveness of the public health response. Additionally, social vulnerabilities and inconsistent public health messaging further exacerbated the pandemic's impact on marginalized communities. This study underscores the critical need for targeted public health strategies that address socio-political and infrastructural disparities to enhance pandemic preparedness and response, emphasizing the importance of equity and effective communication in safeguarding public health."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

The COVID-19 pandemic has cast a stark light on the complexities and challenges of managing a global health crisis, particularly in the United States, which, despite being considered one of the best-prepared nations for a pandemic, experienced significant disparities in outcomes across its diverse states and counties. These disparities highlight the crucial interplay between local health infrastructure, socio-political factors, and the effectiveness of public health responses, underscoring the need for a nuanced understanding of how these elements influence pandemic outcomes.

[@fig-GHS] gives us some context to motivate our research. This map allows us to visualize the Global Health Security (GHS) Index which is designed to assess a country's capabilities to prevent, detect, and respond to significant infectious disease outbreaks. A higher Global Health Security (GHS) Index score indicates a stronger health system capacity for pandemic readiness. The GHS Index assesses the readiness of countries for pandemics and other significant biological threat emergencies, using a comprehensive framework that includes measurements grouped into six broad categories: prevention, detection, response, health system capacity, compliance with international norms, and risk environment. Higher scores suggest that a country has a more robust health system capacity, with better capabilities to prevent, detect, respond to, and mitigate the spread of an epidemic.

@@

Later on, this paper delves into the intricate dynamics between local health systems, socio-political influences, and their collective impact on the disparities observed in COVID-19 outcomes across the United States. By analyzing data from various states and counties, we aim to identify how differences in health infrastructure, political affiliations, and social vulnerabilities contributed to the varied effectiveness of the response to the pandemic. This investigation seeks to fill the gap in understanding the role of these factors in shaping public health outcomes during a pandemic, a critical area of inquiry given the unprecedented scale and impact of COVID-19.

Our research uncovered that political polarization significantly affected adherence to health guidelines, with stark differences in compliance and outcomes between regions with varying political leanings. Furthermore, disparities in healthcare access and the prevalence of social vulnerabilities were closely correlated with the severity of the pandemic's impact, highlighting the exacerbating effect of pre-existing inequalities. These findings underscore the necessity of developing public health strategies that are sensitive to socio-political dynamics and infrastructural disparities, to enhance resilience against future health crises.

This paper is structured to first provide a comprehensive overview of the broader context and motivations behind this study, followed by a detailed examination of the methodologies employed to investigate the research question. The results section presents the key findings, which are then discussed in the context of their implications for public health preparedness and response strategies. The conclusion summarizes the study's contributions to our understanding of the complex factors influencing pandemic outcomes and outlines recommendations for future research and policy development.
